---
title: "Metabolomics data"
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
---

```{r echo=TRUE, message=FALSE, warning=FALSE}
suppressWarnings({
library(dplyr)
library(tidyverse)
library(data.table)
library(caret)
library(FSelector)
library(olsrr)
library(smotefamily)
library(kernlab)
})
```


```{r}
df_meta <- fread('C:\\Users\\sonai\\Omics notebooks\\breast_metabolomics.csv')
```


```{r}
head(df_meta)
```
```{r}
df_meta <- df_meta %>% 
  mutate(Label = as.factor(Label)) 
```

```{r}
head(df_meta)
```


```{r}
str(df_meta)
```

These are the dimensions of the dataframes.
```{r}
dim(df_meta)
```

There are no missing values in any of the dataframes.
```{r}
sum(is.na(df_meta))
```

The labels for the data is very imbalanced.

```{r}
sum(df_meta$Label == 0)
sum(df_meta$Label == 1)
sum(df_meta$Label == 1) / (sum(df_meta$Label == 1)+sum(df_meta$Label == 0))
```


# 1 Feature Selection Methods

# 1.1 Information gain

Information gain is mainly used for continuous target variables, so we won't consider it too much later.

```{r}
ig_weights <- information.gain(Label ~ ., data = df_meta)
ig_weights <- ig_weights[order(ig_weights$attr_importance, decreasing = TRUE),, drop=FALSE]
print(ig_weights)
```

## 1.2 Chi-square test

We select the top 50 features
```{r}
chisq_weights <- chi.squared(Label ~ ., data = df_meta)
chisq_weights <- chisq_weights[order(chisq_weights$attr_importance, decreasing = TRUE),, drop=FALSE]
print(chisq_weights[1:50, , drop=FALSE])
```

## 1.3 Recursive feature elimination

```{r}
x <- df_meta[, -1]
y <- df_meta$Label

control_rfe <- rfeControl(functions = rfFuncs, method = "cv", number = 5)

rfe_results <- rfe(x, y, sizes = c(50), rfeControl = control_rfe)

#Select top 50 features
selected_features_rfe <- predictors(rfe_results)
print(selected_features_rfe)
```

We see that among the top 50 features from both methods, 24 are common in both.

```{r}
top50_features <- rownames(chisq_weights[1:50, , drop = FALSE])

common_features <- intersect(selected_features_rfe, top50_features)

print(common_features)
```
```{r}
diff_features <- setdiff(selected_features_rfe, top50_features)
print(diff_features)
```
# 2. Feature extraction methods

# 2.1. PCA

```{r}
pca_model <- prcomp(x, center = TRUE, scale. = TRUE)

var_explained <- summary(pca_model)$importance[3, ]
n_pc <- min(which(var_explained >= 0.75))
cat("Number of Principle Components to explain 75% variance:", n_pc, "\n")

```
```{r}
pca_data <- data.frame(pca_model$x[, 1:n_pc])
pca_data$Label <- df_meta$Label
```

```{r}
dim(pca_data)
```


# 2.2 Kernel PCA

We use the rbf kernel here.

```{r}
kpca_model <- kpca(~., data=as.data.frame(x), kernel='rbfdot', features=30)
```

```{r}
kpca_features <- rotated(kpca_model)
dim(kpca_features)
```
```{r}
df_kpca <- data.frame(kpca_features, Label = y)
dim(df_kpca)
```

# 3. Classification

We will use random forest as our classification method and the metrics will be the AUC, sensitivity and specificity. We will employ a 5-fold cross validation and do upsampling since the classes are very imbalanced here. We will compare with the features obtained from feature selection and feature extraction.

```{r}
train_control <- trainControl(method = "cv",
                              number = 5,
                              sampling = "up",
                              summaryFunction = twoClassSummary,
                              classProbs = TRUE)
```

## 3.1 Using random forest and chi square features

```{r}
top_chi_features <- rownames(chisq_weights)[1:50]

# Create new dataframe with selected features
data_chi <- df_meta %>%
  select(all_of(top_chi_features), Label)

data_chi <- data_chi %>%
  mutate(Label = factor(Label,
                       levels = c(0, 1),
                       labels = c("Class0", "Class1")))

model_chi <- train(Label ~ ., data = data_chi,
                   method = "rf",
                   metric = "ROC",
                   trControl = train_control)
print(model_chi)
```
## 3.2 Using random forest and rfe features

```{r}
data_rfe <- df_meta %>%
  select(all_of(selected_features_rfe), Label)

data_rfe <- data_rfe %>%
  mutate(Label = factor(Label,
                       levels = c(0, 1),
                       labels = c("Class0", "Class1")))

model_rfe <- train(Label ~ ., data = data_rfe,
                   method = "rf",
                   metric = "ROC",
                   trControl = train_control)
print(model_rfe)

```
## 3.3 Using random forest and pca features

```{r}
pca_data <- pca_data %>%
  mutate(Label = factor(Label,
                       levels = c(0, 1),
                       labels = c("Class0", "Class1")))

model_pca <- train(Label ~ ., data = pca_data,
                   method = "rf",
                   metric = "ROC",
                   trControl = train_control)
print(model_pca)
```
## 3.4 Using random forest and kernel pca features

```{r}
df_kpca <- df_kpca %>%
  mutate(Label = factor(Label,
                       levels = c(0, 1),
                       labels = c("Class0", "Class1")))

model_kpca <- train(Label ~ ., data = df_kpca,
                   method = "rf",
                   metric = "ROC",
                   trControl = train_control)

print(model_kpca)

```


# 4. Feature importance

## 4.1 Feature importance from chi square model

```{r}
importance_chi <- varImp(model_chi)
print(importance_chi)

```


## 4.2 Feature importance from rfe model

```{r}
importance_rfe <- varImp(model_rfe)
print(importance_rfe)

```






